Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler

### 2019-05-01 draft_paper/draft_paper.tex
Reorganize build system to use nix
Start final paper

\documentclass[11pt]{article}

\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}

\usepackage{amsthm}
\usepackage{thmtools}
\declaretheorem[style=plain,numberwithin=section]{theorem}
\declaretheorem[style=plain,sibling=theorem]{proposition}
\declaretheorem[style=plain,sibling=theorem]{lemma}
\declaretheorem[style=plain,sibling=theorem]{corollary}

\declaretheorem[style=definition,numberwithin=section]{definition}
\declaretheorem[style=definition,numberwithin=section]{example}
\declaretheorem[style=definition,numberwithin=section]{remark}

\setlength{\parskip}{1em}
\setlength{\parindent}{0em}
\usepackage[letterpaper,margin=1in]{geometry}

\author{Jake Waksbaum}
\date{\today}
\title{Thesis}
\hypersetup{
 pdfauthor={Jake Waksbaum},
 pdftitle={Thesis},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 26.1 (Org mode 9.1.9)},
 pdflang={English}}

@@

\maketitle
\tableofcontents

@@

The Burrows-Wheeler transform (BWT) is a reversible transformation
that makes a string more amenable to compression by other
methods \cite{bw}. The forwards-transform is straightforward to
describe: for a string of length \(n\), form the \(n \times n\) matrix of
all cyclic shifts of the original string, sort it lexicographically,
and take the last column of the matrix. In addition, we note that
index where the original string appears in the sorted rotation matrix.
As shown in \autoref{fig:bw_ex}, applying the algorithm to the string
\verb|abracadabra!| produces the new string \verb|ard!rcaaabb| and the
index \verb|3|. The tendency of the transformation to group the same
character together is what makes the resulting restring easier to
compress. Although it is not at all obvious, the transformed string,
together with the index, provide enough information to recover the the
original string. BWT is discussed more in \autoref{sec:bwt}.

\begin{figure}
  \centering
  \begin{tabular}{rc}
    0  & abracadabra! \\
    1  & bracadabra!a \\
    2  & racadabra!ab \\
    3  & acadabra!abr \\
    4  & cadabra!abra \\
    5  & adabra!abrac \\
    6  & dabra!abraca \\
    7  & abra!abracad \\
    8  & bra!abracada \\
    9  & ra!abracadab \\
    10 & a!abracadabr \\
    11 & !abracadabra
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{rc}
    0  & !abracadabr\textbf{a} \\
    1  & a!abracadab\textbf{r} \\
    2  & abra!abraca\textbf{d} \\
    \textit{3} & \textit{abracadabra\textbf{!}} \\
    4  & acadabra!ab\textbf{r} \\
    5  & adabra!abra\textbf{c} \\
    6  & bra!abracad\textbf{a} \\
    7  & bracadabra!\textbf{a} \\
    8  & cadabra!abr\textbf{a} \\
    9  & dabra!abrac\textbf{a} \\
    10 & ra!abracada\textbf{b} \\
    11 & racadabra!a\textbf{b}
  \end{tabular}
  \caption{BWT acting on \texttt{abracadabra!}}
  \label{fig:bw_ex}
\end{figure}

Bird and Mu \cite{birdmu,pearls} analyze the Burrows-Wheeler algorithm
in a functional setting by expressing the algorithm as a composition
of many smaller functions. Then, they derive an expression for the
inverse-transform from the specification that it invert the
forwards-transform. In addition to providing an intuitive derivation
of the inverse-transform, this provides the outline of a proof of the
correctness of the Burrows-Wheeler algorithm. We give our version of
their derivation in \autoref{sec:deriv}

Our contribution is a formalization in the Coq proof assistant of the
derivations in \cite{birdmu,pearls}. This is useful in its own right
as a functional specification for a potential verified implementation
of a program that relies on Burrows-Wheeler \cite{appel-func-spec}.
More generally, this provides evidence that Bird's approach of
program calculation, in which programs are derived from their
specifications via algebraic manipulations, is useful for proving the
correctness of algorithms in a machine-checked setting. Bird has
applied this approach to greedy algorithms, the Boyer-Moore algorithm,
the Knuth-Morris-Pratt algoirthm, Sudoku solvers, arithmetic coding,
and the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and
many more problems throughout his career \cite{pearls}. Finally, as a
part of this proof we develop a theory of stable sorts, proving
equivalence between different definitions of sorted lists and stable
sorts, as well as the equivalence of all stable sort algorithms to
each other.

\section{Burrows-Wheeler}
\label{sec:bw}

The Burrows-Wheeler transform was originally introduced as a pass in a
compression algorithm. First the input is transformed with BWT, which
tends to group together characters. The intuition for why this happens
is that all adjacent characters in the output string are followed in
the input string by characters that are adjacent in the sorted first
column. If there is any correlation between a character and the
character preceeding it, bringing together the characters in the first
column will tend to bring together similar characters in the last
column.

@@

\section{The Derivation}
\label{sec:deriv}
In this section, we give an overview of the derivation without delving
into the proofs of the lemmas. We only derive the simples version of
the algorithm, although the derivations in \cite{birdmu,pearls}
provide additional optimizations that bring the final implementation
closer to the imperative implementation in \cite{bw} and other texts.

\subsection{Forwards-Transform}
\label{subsec:ft}
In general, the Burrows-Wheeler transform can be applied to lists of
any type that can be sorted, so we define our functions to operate on
\verb|list A|. We will discuss the sortable constraint in
Section~\autoref{sec:sort_stable}.

First, we define left and right rotations of a list.
\begin{verbatim}
Definition lrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => tl ++ [hd]
  end.

Definition rrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => last l hd :: init l
  end.
\end{verbatim}

Now we can produce the rotation matrix by repeatedly applying \verb|lrot|.
\begin{verbatim}
Fixpoint iter (f : A -> A) (n : nat) (z : A) : list A :=
  match n with
  | O   => []
  | S m => z :: iter f m (f z)
  end.

Definition rots (l : list A) : list (list A) :=
  iter lrot (length l) l.
\end{verbatim}

Finally, we produce the transformed string by taking the last column
of the sorted rotation matrix. \verb|lexsort| sorts the rotation matrix
lexicographically; it's precise implementation is not important.
\begin{verbatim}
Definition bwp (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: _ => List.map (fun x => last x hd) (lexsort (rots l))
  end.
\end{verbatim}

We also need to find the index in the sorted rotation matrix where the
original string appears.
\begin{verbatim}
Fixpoint findIndex (x : A) (ls : list A) : nat :=
  match ls with
  | [] => 0
  | hd :: tl =>
    match x == hd with
    | left _ => 0
    | right _ => S (findIndex x tl)
    end
  end.
\end{verbatim}
\begin{verbatim}
Definition bwn (l : list A) : nat :=
  findIndex l (lexsort (rots l)).
\end{verbatim}

Without proof, we note that we should have for any nonempty
\verb|x : list A| and \verb|d : A|,
\begin{verbatim}
nth (bwn x) (lexsort (rots x)) d = x
\end{verbatim}

\subsection{Inverse-Transform}
\label{subsec:invt}

Now we have to construct a function \verb|unbwt| such that
\begin{verbatim}
unbwt (bwn xs) (bwp xs) = xs.
\end{verbatim}
If we could recreate the entire sorted rotation matrix from
\verb|bwp x|, it woiuld be simple to recover \verb|x| by indexing into
the matrix at \verb|bwn x|. That is, we would like a function
\verb|recreate| such that
\begin{verbatim}
recreate (bwp l) = lexsort (rots l).
\end{verbatim}

Then, we could define \verb|unbwt| as
\begin{verbatim}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{verbatim}
and we would have the correctness of \verb|unbwt| as a result of the
correctness of \verb|bwn|.

\subsubsection{\texttt{recreate}}
We will recreate the sorted rotation matrix column by column, so
\verb|recreate j| will recreate the first \verb|j| columns of the sorted
rotation matrix.
\begin{verbatim}
Definition cols (j : nat) (l : list A) : list A := map (firstn j) l.

@@

\section{Proving the Lemmas}
\label{sec:lemmas}

@@

\section{Sorting \& Stability}
\label{sec:sort_stable}

\subsection{Sorting}
\label{subsec:sorting}
What do we mean by the constraint that \verb|A| is sortable? We say
that a type must be equppied with a total, decidable preorder in order
to be sorted.
\begin{verbatim}
Class Preord (A : Type):=
  { le : A -> A -> Prop;
    le_trans : forall x y z, le x y -> le y z -> le x z;
    le_total : forall x y, le x y \/ le y x;
    le_dec : forall x y, {le x y} + {~le x y};
  }.
\end{verbatim}
The type of a sorting algorithm will be
\begin{verbatim}
sort : forall (A : Type), Preord A -> list A -> list A
\end{verbatim}

A correct sorting algorithm should satisfy
\begin{verbatim}
forall x, Sorted (sort x) /\ Permutation x (sort x)
\end{verbatim}
where we define the \verb|Sorted| relation on lists:
\begin{verbatim}
Inductive Sorted: list A -> Prop :=
| Sorted_nil:
    Sorted nil
| Sorted_cons: forall hd tl,
    (forall x, In x tl -> le hd x) ->
    Sorted tl ->
    Sorted (hd :: tl).
\end{verbatim}

We can also have defined \verb|Sorted| like this
\begin{verbatim}
Inductive LocallySorted : list A -> Prop :=
| LSorted_nil : LocallySorted nil
| LSorted_cons1 a : LocallySorted (a :: nil)
| LSorted_consn a b l :
    LocallySorted (b :: l) -> le a b -> LocallySorted (a :: b :: l).
\end{verbatim}

\begin{theorem}[Sorted-LocallySorted Equivalent]
\begin{verbatim}
forall l, Sorted l <-> LocallySorted l.
\end{verbatim}
\end{theorem}

\begin{proof}
  First we prove the forwards direction by induction on the evidence
  \verb|Sorted l|. For the base case, we have \verb|Lsorted_nil|. In
  the inductive case, we consider when \verb|tl| is empty and when it
  is \verb|b :: tl|. If it is empty, we have \verb|Lsorted_cons1 hd|.
  If it is not, we have \verb|LocallySorted (b :: tl)| by the
  inductive hypothesis. We also have that
  \verb|forall x, In x (b :: tl) -> le hd x)|, and because clearly
  \verb|In b (b :: tl)| we have \verb|le hd b|. This means we can
  apply \verb|Lsorted_consn| and we are done.

  Now we prove the backwards direction by induction on the evidence
  \verb|LocallySorted l|. The nil case is given by \verb|Sorted_nil|,
  and the singleton case is given by the \verb|Sorted_cons| followed
  by \verb|Sorted_nil|, using the fact that \verb|In x []| is
  \verb|False|. In the inductive case we have \verb|Sorted (b :: tl)|
  by the inductive hypothesis, which implies \verb|Sorted tl|. It
  suffices to show \verb|forall x, In x (b :: tl) -> le a x|.
  \verb|Sorted (b :: tl)| implies that
  \verb|forall x, In x tl -> le b x|, and we have that \verb|le a b|,
  so by the transitive property of \verb|le| we are done.
\end{proof}

We could also define sorted based on indexes:
\begin{verbatim}
Definition IndexSorted (l : list A) := forall i j d,
    i <= j < length l ->
    le (nth i l d) (nth j l d).
\end{verbatim}

\begin{theorem}[Sorted-IndexSorted Equivalent]
\begin{verbatim}
forall l, Sorted l <-> IndexSorted l.
\end{verbatim}
\end{theorem}

\begin{theorem}[Permutation-IndexPermutation Equivalent]
  The built-in \verb|Permutation xs ys| relation is equivalent to saying
  that there exists some permutation function \verb|p : 'S_n| such that
\begin{verbatim}
forall i : 'I_n, nth i xs = nth (p i) ys
\end{verbatim}
\end{theorem}

\subsection{Stability}
Any preorder induces a decidable equivalence relation:
\begin{verbatim}
Definition eqv {A} `{Preord A} (x y : A) : Prop := le x y /\ le y x.
\end{verbatim}
A stable sort is a sort that preserves the relative order of
equivalent elements. We can state this naturally in terms of the
permutations. We say that a permutation \verb|p| is stable on a list
\verb|l| if
\begin{verbatim}
forall i j : 'I_n, i < j -> eqv (nth i l) (nth j l) -> p i < p j
\end{verbatim}
Then, a sort is stable if for all \verb|l|, there exists a stable
permutation between \verb|l| and \verb|sort l|.

We can also say that two lists are stably related if
\begin{verbatim}
forall x, filter (eqv_dec x) l = filter (eqv_dec x) l'
\end{verbatim}

\begin{theorem}[PermStable-FilterStable Equivalent]
\begin{verbatim}
forall l,
  PermStable l l' <-> (FilterStable l l' /\ Permutation l l')
\end{verbatim}
\end{theorem}

Stability is important because we can prove that there is only one way
to stably sort any list. This also means that any two correct stable
sorting algorithms must agree always, and can be substituted for one another.

\begin{theorem}[StableUnique]
\begin{verbatim}
forall l l',
  Sorted l' -> Permutation l l' -> Stable l l' -> l = l'
\end{verbatim}
\end{theorem}

If we have a antisymmetric total preorder, also known simply as a
total order, then we have the important result that all sorts are
stable sorts. Antisymmetricity means that
\begin{equation*}
  \forall x y, x \leq y \wedge y \leq x \implies x = y
\end{equation*}
or in terms of equivalence
\begin{equation*}
  \forall x y, x \equiv y \implies x = y
\end{equation*}

\begin{theorem}[OrderStable]
\begin{verbatim}
forall (O : Preorder A) (l l' : A),
  (forall x y, eqv x y -> x = y) ->
  Permutation l l' -> Sorted l -> Sorted l' -> Stable l l'
\end{verbatim}
\end{theorem}

We require that the type \verb|A| be equipped with a decidable total
order. If \verb|A| has a decidable total order, then the lexicographic
preorder on \verb|list A| is also an order. This means that any
sorting algorithm using lexicographic order is stable, including
radix-sort. This is what allows us to conclude that \verb|lexsort| is
equivalent to radix-sort.

\begin{theorem}[RadixSort Correct]
\begin{verbatim}
forall x,
  Sorted (radixsort x) /\ Permutation x (radixsort x)
\end{verbatim}
\end{theorem}

\nocite{*}

\bibliography{thesis}{}

\bibliographystyle{plain}

@@

>>>>

\documentclass[11pt]{thesis}

\title{Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler}
\author{Jake Waksbaum}
\date{May 6, 2019}

@@

\maketitle

@@

The Burrows-Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. The forwards transform is straightforward to describe: for
a string of length \(n\), form the \(n \times n\) rotaion matrix of all
cyclic shifts of the original string, sort it lexicographically, and
take the last column of the matrix. In addition, we note that index
where the original string appears in the sorted rotation matrix. As
shown in \cref{fig:bw_ex}, applying the algorithm to the string
\texttt{abracadabra!} produces the new string \texttt{ard!rcaaabb} and
the index \texttt{3}. The transformation tends to produce runs like
the three \texttt{a}s or the two \texttt{b}s, and it is this feature
that makes the resulting string easier to compress. Although it is not
at all obvious, the transformed string, together with the index,
provide enough information to recover the the original string. BWT is
discussed more in \cref{sec:bwt}.

\begin{figure}
  \centering
  \begin{tt}
  \begin{tabular}{rc}
    0  & abracadabra! \\
    1  & bracadabra!a \\
    2  & racadabra!ab \\
    3  & acadabra!abr \\
    4  & cadabra!abra \\
    5  & adabra!abrac \\
    6  & dabra!abraca \\
    7  & abra!abracad \\
    8  & bra!abracada \\
    9  & ra!abracadab \\
    10 & a!abracadabr \\
    11 & !abracadabra
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{rc}
    0  & !abracadabr\textbf{a} \\
    1  & a!abracadab\textbf{r} \\
    2  & abra!abraca\textbf{d} \\
    \textit{3} & \textit{abracadabra\textbf{!}} \\
    4  & acadabra!ab\textbf{r} \\
    5  & adabra!abra\textbf{c} \\
    6  & bra!abracad\textbf{a} \\
    7  & bracadabra!\textbf{a} \\
    8  & cadabra!abr\textbf{a} \\
    9  & dabra!abrac\textbf{a} \\
    10 & ra!abracada\textbf{b} \\
    11 & racadabra!a\textbf{b}
  \end{tabular}
  \end{tt}
  \caption{BWT acting on \texttt{abracadabra!}}
  \label{fig:bw_ex}
\end{figure}

Bird and Mu \cite{birdmu,pearls} analyze the Burrows-Wheeler algorithm
in a functional setting. They express the algorithm as the composition
of many smaller functions, and then derive an expression for the
inverse transform from the specification that it inverts the forwards
transform. In addition to providing an intuitive derivation of the
inverse transform, this provides the outline of a proof of the
correctness of the Burrows-Wheeler algorithm. We give our version of
their derivation in \cref{sec:forwards_BWT,sec:inverse_BWT}.

Our main contribution is a formalization in the Coq proof assistant of
the BWT, together with a machine-checked proof that our implementation
is correct. Specifically, we implement the functions
$\mathtt{bwp} : \mathtt{list}~A \to \mathtt{list}~A$ and
$\mathtt{bwn} : \mathtt{list}~A \to \mathbb{N}$ to produce the string
and index part of the BWT respectively, and the function
$\mathtt{unbwt} : \mathbb{N} \to \mathtt{list}~A \to \mathtt{list}~A$ to
reverse the transform, and we prove the following theorem:
\begin{theorem}[unbwt\_correct]
  \label{thm:unbwt_correct}
  For all lists $l$,
  \begin{equation*}
    \mathtt{unbwt}~(\mathtt{bwn}~l)~(\mathtt{bwp}~l) = l
  \end{equation*}
\end{theorem}
The implementation of the functions and the outline of the proof are
based on Bird et~al.'s derivations.

The implementation is useful in its own right as a functional
specification for a potential verified implementation of a program
that relies on Burrows-Wheeler \cite{appel-func-spec}. But more
generally, this provides evidence that Bird's approach of program
calculation, in which programs are derived from their specifications
via algebraic manipulations, is useful for proving the correctness of
algorithms in a machine-checked setting. Bird has applied this
approach to greedy algorithms, the Boyer-Moore algorithm, the
Knuth-Morris-Pratt algoirthm, Sudoku solvers, arithmetic coding, and
the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many
more problems \cite{pearls}.

As a part of the proof of \Cref{thm:unbwt_correct}, we implemented
radix-sort and proved its correctness,
\begin{theorem}[radixsort\_correct]
  \label{thm:radixsort_correct}%
  For all lists $l$,
  \begin{equation*}
    \mathtt{Sorted}~(\mathtt{radixsort}~l) \land \mathtt{Permutation}~(\mathtt{radixsort}~l)~l
  \end{equation*}
\end{theorem}
and proved that any two stable sorts are interchangable,
\begin{theorem}[stable\_sort\_unique]
  \label{thm:radixsort_correct}%
  For all functions $f, g : \mathtt{list}~A \to \mathtt{list}~A$,
  \begin{gather*}
    (\forall l,~\mathtt{Sorted}~(f~l) \land \mathtt{StablePermutation}~(f~l)~l) \implies \\
    (\forall l,~\mathtt{Sorted}~(g~l) \land \mathtt{StablePermutation}~(g~l)~l) \implies \\
    \forall l,~f~l = g~l
  \end{gather*}
\end{theorem}
This in turn required developing suitable definitions for
\texttt{Sorted}, \texttt{Permutation}, and \texttt{StablePermutation}.
We explored multiple definitions for each of these properties, and
proved equivalence between them.

The rest of this paper is organize as follows. \Cref{sec:bwt} details
the history of the BWT and its applications. \Cref{sec:ord} defines
what we need to know about a type $A$ to sort a $\mathtt{list}~A$ as a
part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
the implementation of the forwards BWT in Coq, and
\cref{sec:inverse_BWT} derives an implementation of the inverse BWT
via program calculation. \Cref{sec:sorted_perm_stable} provides
multiple definitions of \texttt{Sorted}, \texttt{Permutation}, and
\texttt{StablePermutation}, and proves equivalence between them.
\Cref{sec:sort} implements insertion sort and radixsort, and proves
their correctness. \Cref{sec:BWT_proof} more formally proves the
theorems about the BWT mentioned in previous sections, including
\Cref{thm:unbwt_correct}.

The full source code can be found at
\url{https://github.com/jbaum98/verifed_bzip2/tree/2019-05-06}.

\section{Burrows-Wheeler}
\label{sec:bwt}

The Burrows-Wheeler transform was originally introduced as a pass in a
compression algorithm. First the input is transformed with BWT, which
tends to group together characters. The intuition for why this happens
is that all adjacent characters in the output string are followed in
the input string by characters that are adjacent in the sorted first
column. If there is any correlation between a character and the
character preceding it, bringing together the characters in the first
column will tend to bring together similar characters in the last
column.

@@

\section{Orders and Preorders}
\label{sec:ord}

Although it is commonly demonstrated on strings, the BWT can be
applied to any list whose elements are comparable. To make that
precise, if we have a type $A$ and a relation on that type $\le$, we say
that this forms a preorder on $A$ when $\le$ is reflexive and transitive, that is
$x \le x$ for all $xA$, and
\begin{equation*}
  x \le y \implies y \le z \implies x \le z
\end{equation*}
for all $x,y,z$.

We say that we have a \textit{total} preorder if for any two elements
$x$ and $y$, $x \le y \lor y \le x$.

We say that we have a \textit{decidable} preorder if we have a
decision procedure for determining whether $x \le y$ or $x \not\le y$.

We represent a total, decidable preorder on type with the
\texttt{Preord} typeclass in Coq:
\lstinputlisting[firstline=9, lastline=14]{../theories/BWT/Sorting/Ord.v}
We do not explicitly require reflexivity because it follows from totality.

From $\le$, we can define $x \ge y \coloneqq y \le x$, $x < y \coloneqq
\lnot (y \le x)$, and $x > y \coloneqq \lnot (x \le y)$, and prove
transitivity and reflexivity where appropriate.

We can also define $x \equiv y \coloneqq x \le y \land y \le x$. $\equiv$ is a
decidable equivalence relation, which means it is reflexive,
transitive, and symmetric. The builtin class \texttt{EqDec} is used to
describe types with decidable equivalence relations, so for any
\texttt{Preord} instance we have the \texttt{EqDec} instance
\texttt{Preord\_EqDec} using \texttt{eqv\_dec}:
\lstinputlisting[firstline=116, lastline=120,
  widthgobble=2]{../theories/BWT/Sorting/Ord.v}
In Coq, the notation \texttt{===} is used for the relation of an
\texttt{EqDec} instance and \texttt{==} for the decision procedure of
an \texttt{EqDec} instance.

We say that we have a total, decidable \textit{order} when we have
that for all $x$ and $y$,
\begin{equation*}
  x \equiv y \implies x = y
\end{equation*}
that is, the induced equivalence relation of the preorder ($\equiv$)
implies Leibniz equality ($=$). We use the \texttt{Ord} class in Coq for this:
\lstinputlisting[firstline=175, lastline=176,
  widthgobble=2]{../theories/BWT/Sorting/Ord.v}
For the BWT, we require an \texttt{Ord} instance for $A$, but for other parts of the proof, such as sorting, \texttt{Preord} will suffice.

\section{The Forwards BWT}
\label{sec:forwards_BWT}

First, we define left and right rotations of a list.
\lstinputlisting[widthgobble=2, firstline=17, lastline=22]{../theories/BWT/Lib/List.v}
\lstinputlisting[widthgobble=2, firstline=16, lastline=21]{../theories/BWT/Rotation/Rotation.v}
\lstinputlisting[widthgobble=2, firstline=23, lastline=28]{../theories/BWT/Rotation/Rotation.v}

Now we can produce the rotation matrix by repeatedly applying \texttt{lrot}.
\lstinputlisting[widthgobble=2, firstline=11, lastline=18]{../theories/BWT/Lib/Iterate.v}
\lstinputlisting[widthgobble=2, firstline=17, lastline=18]{../theories/BWT/Rotation/Rots.v}

Finally, we produce the transformed string by taking the last column
of the sorted rotation matrix. \texttt{lexsort} sorts the rotation
matrix lexicographically using the \texttt{Ord} instance; we will
discuss it's implementation in \cref{sec:sort}.
\lstinputlisting[firstline=31,
  lastline=35]{../theories/BWT/BurrowsWheeler.v}
Note that in order to ensure that $\mathtt{last}$ is total, it takes
an extra argument to return in the case of an empty list.

We also need to find the index in the sorted rotation matrix where the
original string appears.
\lstinputlisting[firstline=12, lastline=19]{../theories/BWT/Lib/FindIndex.v}
\lstinputlisting[firstline=73, lastline=74]{../theories/BWT/BurrowsWheeler.v}

\section{The Inverse BWT}
\label{sec:inverse_BWT}

Now we have to construct a function \texttt{unbwt} such that for all $l$,
\begin{equation}
  \mathtt{unbwt}~(\mathtt{bwn}~l)~(\mathtt{bwp}~xs) = l
  \label{eq:unbwt}
\end{equation}
If we could recreate the entire sorted rotation matrix from
$\mathtt{bwp}~l$, it would be simple to recover $x$ by indexing into
the matrix at $\mathtt{bwn}~x$.

More precisely, suppose we had a function function \texttt{recreate}
such that
\begin{equation}
  \mathtt{recreate}~(\mathtt{bwp}~l) = \mathtt{lexsort}~(\mathtt{rots}~l)
  \label{eq:recreate}
\end{equation}
Then, we could define \texttt{unbwt} as
\begin{lstlisting}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{lstlisting}
and \Cref{eq:unbwt} would follow directly from the following theorem:
\begin{theorem}[bwn\_correct]
  For every non-empty list $l$,
  \begin{equation*}
    (\mathtt{lexsort}~(\mathtt{rots}~l))_{(\mathtt{bwn}~l)} = l
  \end{equation*}
\end{theorem}

How, then, do we implement \texttt{recreate}? We will recreate the
sorted rotation matrix column by column, so \verb|recreate j| will
recreate the first \verb|j| columns of the sorted rotation matrix.
\begin{verbatim}
Definition cols (j : nat) (l : list A) : list A := map (firstn j) l.

@@

\section{\texttt{Sorted}, \texttt{Permutation}, and \texttt{StablePermutation}}
\label{sec:sorted_perm_stable}

@@

\section{Insertion Sort, Radixsort, and \texttt{lexsort}}
\label{sec:sort}

\section{Proving BWT Correct}
\label{sec:BWT_proof}

\printbibliography{}

@@

We prove this using a few lemmas about \texttt{findIndex},
\texttt{lexsort}, and \texttt{rots}.

\begin{lemma}[findIndex\_correct]
  For all elements $x$, lists $l$,
  \begin{equation*}
    (\exists y \in l,~x \equiv y) \implies l_{(\mathtt{findIndex}~x~xs)} \equiv x
  \end{equation*}
  \begin{proof}
    This follows by induction on $xs$: At every step, either the
    current element is equivalent to $x$ and we return $0$, or there
    must be such a $y$ in the rest of the list.
  \end{proof}
\end{lemma}

\begin{lemma}[orig\_in\_sorted\_rots]
  For every non-empty list $l$,
  \begin{equation*}
    \exists y \in \mathtt{lexsort}~(\mathtt{rots}~l),~x = y
  \end{equation*}
  \begin{proof}
    Clearly $l$ is in $\mathtt{rots}~l$ at index 0, and because
    \texttt{lexsort} produces a permutation of its input, $l$ must be
    in $\mathtt{lexsort}~(\mathtt{rots}~l)$ as well.
  \end{proof}
\end{lemma}

\begin{theorem}[bwn\_correct]
  For every non-empty list $xs$,
  \begin{equation*}
    (\mathtt{lexsort}~(\mathtt{rots}~xs))_{(\mathtt{bwn}~xs)} = xs
  \end{equation*}
  \begin{proof}
    This follows from \textsf{findIndex\_correct} if we can show that
    $\exists y \in xs,~x = y$, which we have by \textsf{orig\_in\_sorted\_rots}.
  \end{proof}
\end{theorem}
